INGRID: Creating Languages in MPS from ANTLR Grammars

### 2018-05-09 chapters/evaluation.tex
minor changes
shortening

However, first we must emphasize that MPS languages automatically produced by the \textsc{Ingrid} method, are not ready-to-use full-fledged MPS counterparts of the original input languages.
The structure of a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects have to be manually tweaked (e.g., the Editor) or defined from scratch by the end user.
\textsc{Ingrid} also does not yet support advanced features of MPS, such as type checking.

For each of the three languages (JSON, JavaScript, C\#), we provide (1) its MPS definition in the form that incorporates all tweaks performed manually by authors of this paper and (2) the adjusted ANTLR v4 grammar used as input for \textsc{Ingrid}.
%, both in the repository that contains also our implementation.

\paragraph{JSON.}
The least amount of manual adjusting after the import into MPS was needed in case of the JSON language~\cite{ref:JSON}, because it is the simplest language from all that we used for our experiments.
Specifically, the first author spent less than 20 minutes in order to get a language that is ready to use.

\paragraph{JavaScript.}
In the case of JavaScript, which is an example of a widely-used complex general purpose programming language, automated generating of the language definition in MPS from the ANTLR grammar and subsequent manual adjusting was done in less than one hour.
%\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}

We are aware of other projects that aim to create a manual port of JavaScript into MPS.
For example, there is ECMAScript4MPS~\cite{ref:ECMA4MPS} developed by the author of the PE4MPS project~\cite{ref:PE4MPS} that we described in Section~\ref{sect:RELATED}.

The main advantage of \textsc{Ingrid} over ECMAScript4MPS is that, despite its current limitations, \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful --- for example, the language designer has to define actions that improve editing experience by automatically transforming the AST in the background.
\textsc{Ingrid} achieved a very good result, when compared to PE4MPS, especially in the case of language structure, concept aliases, and support for auto-completion.
The Structure aspect generated by \textsc{Ingrid} is very similar to that of the ECMAScript4MPS project, which was created manually over a large number of hours.

\paragraph{C\#.}
Another very complex programming language that we used for experiments is C\#~\cite{ref:CSHARP}.
Before we could run \textsc{Ingrid}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure (hierarchy of concepts) that can be practically used in MPS just with minor alterations.
Necessary adjustments include removal of intermediary rules such as \antlrparserrule{statement\_list} : \antlrparserrule{statement*}~\antlrliteral{;} by inlining the right-hand side.
The Editor aspect was modified afterwards, but only for some of the concepts. 
Roughly one hour of manual effort was needed in total.
Nevertheless, additional modifications of the grammar and selected aspects of the MPS language are still needed.
A great flexibility and complexity of the C\# language is the main reason behind all the necessary changes to the language definition in MPS and to the ANTLR grammar.
The definition is quite large, involving more than 800 concepts.
%, and therefore compilation (building) of the language in MPS takes a rather long time.

\paragraph{Other languages.}
We also tried to apply \textsc{Ingrid} on few other languages, such as Python and Ruby.
Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with \textsc{Ingrid}.
For example, the structure and hierarchy of rules in the Python grammar are quite different from JavaScript or JSON, and consequently the language definition created by \textsc{Ingrid} is rather badly organized.
We plan to address this limitation in the future and extend \textsc{Ingrid} such that very few modifications of the grammar would be necessary even for these languages.

\paragraph{General observations.}
The main overall benefit of the \textsc{Ingrid} method is partial automation.
Most of the languages discussed above are quite complex regarding their structure and syntax variety.
Therefore, completely manual definition would be a very time-consuming and error-prone process.
Fully automated generation of the Structure aspect is where \textsc{Ingrid} spares the user from many hours of tedious and sometimes quite challenging work.

On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, an order of magnitude less time than we initially expected.
The first author spent between 20 and 60 minutes of work on each language, when adjusting the result of automated generation into a more useful and readable form.
MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing designers to produce really useful languages very fast.
%All of this justifies our decisions that we have described in Section~\ref{sect:EDITORDEF} and Section~\ref{sect:TEXTGENDEF}, proving that our approach is quite practical.
Despite that, we also tried to design some automated heuristics, but so far all yield rather mediocre results when compared to what human users can achieve efficiently instead.

>>>>

However, first we must emphasize that MPS languages automatically produced by \textsc{Ingrid}, are not ready-to-use full-fledged MPS counterparts of the original input languages.
The structure of a generated MPS language fully corresponds to the respective ANTLR grammar, but its other aspects have to be manually tweaked (e.g., the Editor) or defined from scratch by the end user.
\textsc{Ingrid} also does not yet support advanced features of MPS, such as type checking.

\noindent\textbf{JSON.}
The least amount of manual adjusting after the import into MPS was needed in case of the JSON language~\cite{ref:JSON}, because it is the simplest language from all that we used for our experiments.
Specifically, the first author spent less than 20 minutes in order to get a language that is ready to use.

\noindent\textbf{JavaScript.}
In the case of JavaScript, automated generating of the language definition in MPS from the ANTLR grammar and subsequent manual adjusting was done in less than one hour.
The current version of \textsc{Ingrid} fully automatically produces the definition of JavaScript in MPS that needs just minor adjustments to be really useful --- for example, the language designer has to define actions that improve editing experience by automatically transforming the AST in the background.

\noindent\textbf{C\#.}
Before we could run \textsc{Ingrid} on C\#~\cite{ref:CSHARP}, we had to manually adjust the ANTLR grammar of C\# in order to ensure that \textsc{Ingrid} produces at least a reasonable structure (hierarchy of concepts) that can be practically used in MPS just with minor alterations.
Necessary adjustments include removal of intermediary rules such as \antlrparserrule{statement\_list} : \antlrparserrule{statement*}~\antlrliteral{;} by inlining the right-hand side.
The Editor aspect was modified afterwards, but only for some of the concepts. 
Roughly one hour of manual effort was needed in total.
Nevertheless, additional modifications of the grammar and selected aspects of the MPS language are still needed.
A great flexibility and complexity of the C\# language is the main reason behind all the necessary changes to the language definition in MPS and to the ANTLR grammar.
The definition is quite large, involving more than 800 concepts.

\noindent\textbf{Other languages.}
We also tried to apply \textsc{Ingrid} on few other languages, such as Python and Ruby.
Results are mixed because ANTLR grammars of these languages are written in a style that is not fully compatible with \textsc{Ingrid}.

\noindent\textbf{General observations.}
The main overall benefit of the \textsc{Ingrid} method is partial automation.
Most of the languages discussed above are quite complex regarding their structure and syntax variety.
Therefore, completely manual definition would be a very time-consuming and error-prone process.
Fully automated generation of the Structure aspect is where \textsc{Ingrid} spares the user from many hours of tedious and sometimes quite challenging work.

On the other hand, our experiments with complex languages show that, in the case of the Editor and TextGen aspects, manual adjustment (e.g., adding line breaks and indentation) is a very effective approach that takes only a short time --- in particular, an order of magnitude less time than we initially expected.
MPS IDE provides good support for efficient tweaking of the code layout in Editor and TextGen, allowing designers to produce really useful languages very fast.
Despite that, we also tried to design some automated heuristics, but so far all yield rather mediocre results when compared to what human users can achieve efficiently instead.

Regarding future changes to the ANTLR grammar of a subject language, it is certainly possible to directly update the corresponding MPS definition, but in some cases it may be easier to run again the whole import process and overwrite the previous definition with a new one.

### 2018-05-09 chapters/futurework.tex
shortening

\section{Conclusion and Future Work}

Our contribution is the \textsc{Ingrid} method for constructing language definitions in MPS from grammars in the ANTLR notation.
The method is only partially automated, but nevertheless it greatly reduces the amount of manual work that developers of MPS languages have to perform.
Although we discussed all challenges, details of our solution and general observations just in the context of ANTLR grammars and JetBrains MPS, we believe they are more general --- in particular, relevant to everyone tackling a similar problem in the field of DSLs and language workbenches.

In the future, we would like to add support for generating other aspects of MPS languages and increase the level of automation.
Our primary research goal is to investigate the usage of automated (machine) learning for the purpose of generating editors that support nice and readable code layout.
First, we would have to collect a set of input source code files to be used for training. This step might involve a user study designed in order to identify examples of layout that a majority considers as nice and readable.
The information about code layout, once it is available, will be then leveraged and used also to improve TextGen.
Another subject for future work is automated construction of parser-based tools that could be used for translating source code in plain-text to instances of MPS languages.

>>>>

\section{Future Work}

In the future, we would like to add support for generating other aspects of MPS languages and to increase the level of automation.
Our primary research goal is to investigate the usage of automated (machine) learning for the purpose of generating editors that support nice and readable code layout.
First, we would have to collect a set of input source code files to be used for training. This step might involve a user study designed in order to identify examples of layout that a majority considers as nice and readable.
The information about code layout, once it is available, will be then leveraged and used also to improve TextGen.
Another subject for future work is automated construction of parser-based tools that could be used for translating source code in plain-text to instances of MPS languages.

### 2018-05-09 chapters/introduction.tex
shortening

In the next section, we provide (i) the necessary information about the MPS infrastructure for language definition and (ii) a brief overview of projects with similar goals as \textsc{Ingrid}.
Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution.
We also discuss our experience with application of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\# (Section~\ref{sect:EVAL}), and then we conclude.

>>>>

In the next section, we provide (i) the necessary information about the MPS infrastructure for language definition and (ii) a brief overview of projects with similar goals as \textsc{Ingrid}.
Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution.
We also discuss our experience with application of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\# (Section~\ref{sect:EVAL}), and then we outline our plans for the future work.

### 2018-05-09 paper.tex
minor changes
shortening

\documentclass[sigplan,10pt,preprint]{acmart}

@@

\setcounter{tocdepth}{3}

\settopmatter{printfolios=true}

@@

% TODO
%\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
%\acmYear{2017}
%\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
%\acmDOI{10.1145/nnnnnnn.nnnnnnn}

% TODO
%\setcopyright{acmlicensed}

@@

\input{chapters/introduction}
\input{chapters/background}
\input{chapters/language_definition}
\input{chapters/evaluation}
\input{chapters/conclusion}

@@

\bibliographystyle{abbrvnat}

@@

\bibitem{ref:MPSBOOK} F. Campagne. The MPS Language Workbench, Volume I. CreateSpace Independent Publishing Platform, 2014.

\bibitem{ref:BISONFLEX} J. Levine. Flex \& Bison: Text Processing Tools. O'Reilly Media, 2009.

\bibitem{ref:KV10} L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010, ACM.

\bibitem{ref:ANTLRBOOK} T. Parr. The Definitive ANTLR 4 Reference. Pragmatic Bookshelf, 2013.

\bibitem{ref:MBEDDR} M. Voelter, D. Ratiu, B. Schaetz, and B. Kolb. mbeddr: an Extensible C-based Programming Language and IDE for Embedded Systems. SPLASH 2012, ACM.

\bibitem{ref:VSB14} M. Voelter, J. Siegmund, T. Berger, and B. Kolb. Towards User-Friendly Projectional Editors. SLE 2014, LNCS, 8706.

\bibitem{ref:VWK15} M. Voelter, J. Warmer, and B. Kolb. Projecting a Modular Future. IEEE Software, 32(5), 2015.

@@

\bibitem{ref:ANTLR2MPS} F. Campagne. The ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR_MPS}

@@

\bibitem{ref:ECMA4MPS} M. Lombardo. The ECMAScript4MPS project, \url{https://github.com/mar9000/ecmascript4mps}

\bibitem{ref:PE4MPS} M. Lombardo. The PE4MPS project, \url{https://github.com/mar9000/pe4mps}

\bibitem{ref:PE} M. Lombardo. PE grammars, \url{https://github.com/mar9000/pe}

@@

>>>>

\documentclass[sigconf]{acmart}

@@

% TODO
% copyright
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}

% TODO
%\acmDOI{10.475/123_4}

% TODO
%\acmISBN{123-4567-24-567/08/06}

% TODO
%\acmConference[WOODSTOCK'97]{ACM Woodstock conference}{July 1997}{El Paso, Texas USA}
%\acmYear{2018}
%\copyrightyear{2018}

% TODO
%\acmArticle{4}
%\acmPrice{15.00}

@@

@@

\input{chapters/introduction}
\input{chapters/background}
\input{chapters/language_definition}
\input{chapters/evaluation}
\input{chapters/futurework}

@@

\bibliographystyle{ACM-Reference-Format}

@@

\bibitem{ref:MPSBOOK} Fabien Campagne. 2014. The MPS Language Workbench, Volume I. CreateSpace Independent Publishing Platform.

\bibitem{ref:DHL80} Veronique Donzeau-Gouge, Gerard Huet, Bernard Lang, and Gilles Kahn. 1980. Programming Environments Based on Structured Editors: The Mentor Experience. Research Report RR-0026, INRIA.

\bibitem{ref:BISONFLEX} John Levine. 2009. Flex \& Bison: Text Processing Tools. O'Reilly Media.

\bibitem{ref:KV10} Lennart C.L. Kats and Eelco Visser. 2010. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. In Proc. of OOPSLA 2010, ACM.

\bibitem{ref:ANTLRBOOK} Terence Parr. 2013. The Definitive ANTLR 4 Reference. Pragmatic Bookshelf.

\bibitem{ref:MBEDDR} Markus Voelter, Daniel Ratiu, Bernhard Schaetz, and Bernd Kolb. 2012. mbeddr: an Extensible C-based Programming Language and IDE for Embedded Systems. In Proc. of SPLASH 2012, ACM.

\bibitem{ref:VSB14} Markus Voelter, Janet Siegmund, Thorsten Berger, and Bernd Kolb. 2014. Towards User-Friendly Projectional Editors. In Proc. of SLE 2014, LNCS, 8706.

\bibitem{ref:VWK15} Markus Voelter, Jos Warmer, and Bernd Kolb. 2015. Projecting a Modular Future. IEEE Software, 32(5).

\bibitem{ref:TRFULL} Premysl Vysoky, Pavel Parizek, and Vaclav Pech. 2018. INGRID: Creating Languages in MPS from ANTLR Grammars (Full version). Tech. Report No. D3S-TR-2018-01, Dep. of Distributed and Dependable Systems, Charles University.

@@

\bibitem{ref:ANTLR2MPS} Fabien Campagne. The ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR_MPS}

@@

\bibitem{ref:PE4MPS} Marco Lombardo. The PE4MPS project, \url{https://github.com/mar9000/pe4mps}

\bibitem{ref:PE} Marco Lombardo. PE grammars, \url{https://github.com/mar9000/pe}

@@

\input{chapters/appendix}
